\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-07-06 eg6-letters/NIM/EG6_RTPC_NIM.tex

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at the Jefferson 
Laboratory to track low-energy nuclear recoils for the purpose of measuring
exclusive nuclear reactions, such as coherent Deeply Virtual Compton Scattering
and coherent meson production off $^4$He. In such processes, the $^4$He nucleus
remains intact in the final state, however the CEBAF Large
Acceptance Spectrometer (CLAS) cannot track the low energy $\alpha$ particles. In
2009, we carried out measurements using the CLAS spectrometer supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as 12~MeV for $^4$He. This article discusses the design,
principle of operation, calibration methods and the performances of this RTPC.
\end{abstract}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For exemple, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nuclei) necessitates an upgrade of the photon detection system. 
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 
experiment in 2005~\cite{FX}. The IC is constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in RTPC. The CLAS detector supplemented with IC and RTPC was used 
during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 2005~\cite{BONUS-NIM}. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib}. The tracking algorithm is discussed in section
\ref{sec_tracking}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{fig_2017/RTPC_1_all.pdf}
\caption{Schematic drawing of CLAS RTPC in a plane perpendicular to the beam 
direction. See text for description of the elements.} \label{fig:RTPC_1_4}
\end{figure}

@@

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.55]{fig/PADs.png}
   \caption[]{A schematic representation of a part of the readout system.  The 
   shaded sixteen pads are a group of pads that are connected to the same 
pre-amplifier.} \label{fig:PADs}
\end{figure}

The RTPC electron collection system has 3200 readout pads. These elements are
parallel to the GEMs, 69 mm from the central axis.
Figure~\ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width. The shift between the rows permits the reduction of aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads were grouped on 16 channel pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and drive the 6~m long ribbon 
cable that connects to it.

@@

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, and tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the signal 
digitization and elaboration part is implemented in the digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A readout 
control unit (RCU) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs with communication 
between them performed through a 
custom back-plane, implementing a low-voltage signal bus. The RCU communicates 
with the CLAS data acquisition (DAQ) system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC) used to interface with 
the CLAS DAQ system. An Ethernet link is also present, for slow-controls and 
monitoring.

The triggering and data-readout scheme is as follows. When a Level-1 (L1) trigger 
is received by a FEC-board, a programmable number of samples $N_0$ is digitized 
and stored in a data memory, and hence processed by the digital circuit. Upon 
arrival of Level-2 trigger, the latest processed samples are saved in a 
multi-event memory, individual to each-channel, ready for readout by the RCU.  
In our setup, the RCU receives the trigger signal from the CLAS Trigger 
Supervisor system, and distributes both L1 and L2 signals to all the FECs in 
the crate, with a definite L1-L2 trigger latency. Such latency is programmable, 
and has to be long enough to accommodate the digitization of the $N_0$ samples 
per channel.  In the current implementation, $N_0=100$ samples/channel are 
acquired for each trigger, operating ALTRO at the reduced sampling frequency of 
$10$~MHz. The L1-L2 latency is thus fixed to 15 $\mu$s.

The trigger signal also initiates RCU readout operation from FEC boards. All 
the measured samples from active channels are reported, together with a channel 
identifier and a time-stamp, to the RCU application, that in turns sends them 
on to the main event builder. During data reconstruction, the acquired samples are 
processed to obtain, for each readout pad, the accumulated charge (ADC) and the 
pulse time (T). Since pulse time was obtained as the time-stamp of the first 
sample above threshold, the resolution is equivalent to the ALTRO sampling 
time of 100~ns.

In order to reduce the data size, ALTRO is operated in zero-suppression mode: 
only samples above a programmable threshold are stored in the ALTRO on-board 
memory and then written to tape. A glitch-filter permits to reject spurious 
pulses due to noise, by requiring the presence of a minimum number of samples 
over threshold $N_{MINSEQ}$ to validate a sample. To properly reconstruct 
the signal shape, $N_{PRE}$ samples before threshold-crossing and $N_{POST}$ 
samples after the signal returns below threshold are saved. In the present 
configuration, $N_{MINSEQ}=N_{PRE}=N_{POST}=3$, while threshold is set just 
above the noise level.

In order to read all the detector readout pads, four FEC crates are used, 
each equipped with 6 boards, plus a RCU. A schematic of the readout system, 
for a single crate, is reported in Fig.~\ref{fig:ReadoutScheme}. This 
configuration permits to reduce the dead time associated to FEC readout 
operations and to scale linearly with the number of boards in the crate. During the 
2009 run, the system was successfully operated with a DAQ rate of 3.1 kHz, with 
a live time of $70 \%$.

@@

The timing information collected for each signal above threshold
is used to infer the origin of the ionization electrons and 
then the trajectory of the detected particle. Going from a collection of 
times to a momentum measurement based on the curvature of a track 
requires a good knowledge of the drift speed 
and drift paths followed by the electrons released in the gas. The recorded ADCs give 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enable the particle 
identification.

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, with each time an increasing number of events 
properly reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, the cylindrical symmetry insures that the maximum distance is the same for all 
tracks. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

To measure the drift speed, we use the time profile of all hits in the 
chamber shown in Fig.~\ref{fig:TDC_profile}. We can clearly observe the 
dropping edge expected from geometrical considerations. We define a value $T_{Max/2}$ at 
which the dropping edge passes half the maximum number of hits in the 
histogram. This value was measured in bins along the 200~mm RTPC's length to take into 
account variations in the electric and magnetic field in the RTPC (see 
Fig.~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.42]{fig_2017/hits_time_profile.pdf}
   \caption[]{Time distribution of the collected hits in one experimental run. Trigger
              time is defined as T=15, the time unit is the ALTRO sampling time of 100~ns.} 
   \label{fig:TDC_profile}
\end{figure}

@@

Due to the non perfect experimental conditions, in particular possible contamination 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around run number 
61600. While our gas system was kept slightly over 
atmospheric pressure to avoid contamination from air or other external gases, it 
is likely that this leak was the source of modification of the gas mixture.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe significant change in 
the drift speed before and after run 61600, while variations within these periods 
are ~2\%. These variations of the drift speed were accounted in the track reconstruction

\begin{figure*}[tb]
\centering
\includegraphics[width=16.5cm]{fig_2017/Drift_run_number.pdf}
\caption{$T_{max/2}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculate them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

Based on the kinematics of the electrons in the calibration data, 
we generate the helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then 
we compare the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. 
Because of the magnetic field, the drift paths are not linear in the RTPC. So 
to perform the extraction, we make a first approximation with a linear 
dependence between the radius of emission and the time of detection, and then 
refine our result. As it happens, the curvature is minimal and this process 
converges already on the second iteration.

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) is extracted as a function of time. 
In Fig.~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift path is easily identified and eventually fitted for 
implementation in our reconstruction codes.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{fig_2017/FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution in the second pass for tracks
near 35 mm longitudinal position along the RTPC. The black line represents 
the final drift paths in this slice.}
\label{fig:DELTA_PHI_TDC}
\end{figure}

@@

To calibrate the gains, we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charge along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correction factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position.

@@

\begin{figure*}[tb]
\centering
\includegraphics[scale=0.73]{fig_2017/f_dedx_p_exp_2nd.pdf}
\caption{$\small{\frac{dE}{dX}}$ vs. $p/q$ distribution for the left half of 
   the RTPC (left) and for the right half (right) after gain calibration.  
   The lines are theoretical expectations from Bethe-Bloch formula for $^4$He 
   (red), $^3$He (black), $^3$H (blue), $^2$H (pink) and protons (grey).}
\label{fig:dedx_p_exp_2nd}
\end{figure*}

@@

\subsection{Track Fitting}\label{sec_rec}
We start with reconstructing the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit the chains with a helix if they have a minimum of 
10 hits. We then eliminate from the chain the hits that are 5 mm or farther from the fit
as they are not likely part of the same track.
This new reduced chain is used for a second and final helix fit.

@@

\subsection{Resolution}
Elastic scattering was used to estimate the tracking resolution of the RTPC 
based on the residual between the expected and measured $^4$He tracks.  The RTPC
resolutions, after removing contributions from the electron, are shown in Table
~\ref{tab:reso}, and are very similar for the two halves of the RTPC.  Note, the
$\theta$- and $z$-resolutions are highly correlated.
\begin{table}[htbp]
\begin{center}
\begin{tabular}{|l|cccc|}
  \hline
& $\sigma_{z}$ &  $\sigma_{\theta}$ & $\sigma_{\phi}$ & $\sigma_{p}$\\
\hline
Left &  5.3 mm & 3.8$^{\circ}$ & 1.9$^{\circ}$ & 9$\%$ \\
Right & 6.5 mm & 4.0$^{\circ}$ & 1.9$^{\circ}$ & 8$\%$\\
\hline
\end{tabular}
\caption{The resolutions of the two modules of the RTPC for $z$-vertex, polar and azimuthal angles, and momentum.}
\label{tab:reso}
\end{center}
\end{table}

\subsection{Efficiency}
We measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two halves of the detector. We observe that the left and 
the right modules have similar efficiencies except near the upstream target 
window. This difference is due to the large number of dead channels concentrated
in this part of the left half of the detector

>>>>

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at the Jefferson 
Laboratory to track low-energy nuclear recoils to measure
exclusive nuclear reactions, such as coherent deeply virtual Compton scattering
and coherent meson production off $^4$He. In
2009, we carried out these measurements using the CEBAF Large
Acceptance Spectrometer (CLAS) supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as 12~MeV for $^4$He. This article discusses the design,
principle of operation, calibration methods and performances of this RTPC.
\end{abstract}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nuclei) necessitates an upgrade of the photon detection system. 
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 
experiment in 2005~\cite{FX}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in the RTPC. The CLAS detector supplemented with both IC and RTPC was used 
during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 
2005~\cite{BONUS-NIM}. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib} and the tracking algorithm in section
\ref{sec_tracking}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{fig_2017/RTPC_1_all.pdf}
\caption{Schematic drawing of the CLAS RTPC in a plane perpendicular to the beam 
direction. See text for description of the elements.} \label{fig:RTPC_1_4}
\end{figure}

@@

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.55]{fig/PADs.png}
   \caption[]{A schematic representation of the readout pads. The 
   shaded sixteen pads are a group of pads that are connected to the same 
pre-amplifier.} \label{fig:PADs}
\end{figure}

The RTPC electron collection system had 3200 readout pads. These elements were
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} illustrates the configuration of the 5 by 4.45 mm pads,
where the shift between the rows was implemented to reduce aliasing. Each half of the 
RTPC had 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads were grouped to 16 channels pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6-m long ribbon 
cable that connects to it.

@@

The readout system, similar to the original BoNuS RTPC system~\cite{BONUS-NIM}, 
was based on the front end electronic boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. 
The Front End Card (FEC) 
hosted the analogue receiver circuit and the subsequent amplification and 
digitization stage, based on the PASA~\cite{ALICE-PASA} and ALTRO~\cite{ALICE-ALTRO} 
ASICS. A Readout Control Unit (RCU) board was used to distribute the trigger signal 
to FECs and for data readout, with each RCU handling up to 25 FECs. Communication 
between readout controller application (ROC) and FECs was performed trough a custom 
backplane, implementing a low-voltage signal bus. The RCU communicated with the 
CLAS DAQ system trough a 200 MB/s optical link, connected to a data acquisition 
PC hosting a ReadOut Receiver Card (RORC). This PC hosted the ROC used to interface 
with the CLAS DAQ system. An Ethernet link was also present, for slow-controls and 
monitoring.

The standard CLAS electron trigger was used to initiate the RTPC readout. For each 
event, 100 samples/channel were digitized and processed by the ALTRO ASIC. In order 
to reduce the data size, ALTRO was operated in zero-suppression mode, keeping only 
samples from channels above a programmable threshold set just above the noise level. 
Samples were pre-processed 
by a glitch-filter to reject noise and spurious pulses. To properly reconstruct the 
signal shape, $N_{PRE}=3$ samples before threshold-crossing and $N_{POST}=3$ samples 
after the signal returns below threshold were saved.

The trigger signal also initiates RCU readout operation from FEC boards. All the 
measured samples from active channels are reported, together with a channel 
identifier and a timestamp, to the ROC application, that in turns sends them on 
the main event builder. Readout from FECs occurs in ``block-transfer'' mode, 
making use of the ALTRO internal multi-event buffer. This feature could be 
exploited thanks to the new RCU boards developed for the RTPC readout 
system\footnote{In the original BONUS system, the U2F readout controller had 
no multi-event capability and used a USB connection rather than fiber 
optics for readout. These factors limited the maximum readout rate from the 
BONUS detector to about 500 Hz.}. In order to read all the detector readout 
pads, four FEC crates were used, each equipped with 6 boards, plus a ROC. A 
schematic of the readout system, for a single crate, is reported in 
figure~\ref{fig:ReadoutScheme}. This configuration permitted to reduce the dead 
time associated to FEC readout operations, which is scaling linearly with the number of 
boards in the crate. During the 2009 run, the system was successfully operated 
with a DAQ rate of 3.1 kHz and a live time of $70 \%$, for a luminosity of 
about $10^{34}$~cm$^{-2}$.s$^{-1}$ and a beam energy of 6.067 GeV.

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100 ns.

@@

The timing information collected for each signal above threshold
was used to infer the origin of the ionization electrons and 
then the trajectory of the detected particle. The recorded ADCs were used to reconstruct 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enabled the particle 
identification.

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, with each time an increasing number of tracks 
reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, the cylindrical symmetry insures that this maximum drift distance is the same for all 
tracks. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can therefore infer the drift 
speed of the electrons in the RTPC.\\

@@

To measure the maximum drift time, we used the time profile of hits from identified tracks 
shown in Fig.~\ref{fig:TDC_profile}. We can clearly observe the 
dropping edge expected from geometrical considerations. We define a value $T_{Max/2}$ at 
which the dropping edge passes half the maximum number of hits in the 
histogram. This value was measured in bins along the 200~mm RTPC's length to take into 
account variations in the electric and magnetic field in the RTPC (see 
Fig.~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.42]{fig_2017/hits_time_profile.pdf}
   \caption[]{Time distribution of the hits associated with a track in one experimental run. Trigger
              time is defined as T=15, the time unit is the ALTRO sampling time of 100~ns.} 
   \label{fig:TDC_profile}
\end{figure}

@@

Due to the non perfect experimental conditions, in particular possible contamination 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around run number 
61600. While our gas system was kept slightly over 
atmospheric pressure to limit contamination from air or other external gases, it 
is likely that this leak was the source of modification of the drift speed.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe significant change in 
the drift speed before and after run 61600, while variations within these periods 
are ~2\%.

\begin{figure*}[tb]
\centering
\includegraphics[width=15.5cm]{fig_2017/Drift_run_number.pdf}
\caption{$T_{max/2}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We this calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

Based on the kinematics of the electrons in the calibration data, 
we generated the helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then,
we compared the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. To perform this drift path extraction, 
we made a first approximation assuming a linear dependence between the radius 
of emission of the charge and its time of detection, and then refined our 
result. Indeed, because of the magnetic field, the drift paths are not linear 
in the RTPC, but the curvature was minimal and this process converged already 
on the second iteration.

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) was extracted as a function of time. 
In Fig.~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift path is easily identified and eventually fitted for 
implementation in our reconstruction codes.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{fig_2017/FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution for tracks
from one bin in longitudinal position along the RTPC. The black line represents 
the final drift paths in this bin.}
\label{fig:DELTA_PHI_TDC}
\end{figure}

@@

To calibrate the gains, we compared the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charge along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains were refined using correction factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compared it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position.

@@

\begin{figure*}[tb]
\centering
\includegraphics[scale=0.73]{fig_2017/f_dedx_p_exp_2nd.pdf}
\caption{$\small{\frac{dE}{dX}}$ vs. $p/q$ distributions for the left (on the 
   left) and for the right (on the right) half of the RTPC after gain 
   calibration. The lines are theoretical expectations from the Bethe-Bloch 
   formula for $^4$He 
   (red), $^3$He (black), $^3$H (blue), $^2$H (pink) and protons (gray).}
\label{fig:dedx_p_exp_2nd}
\end{figure*}

@@

\subsection{Track Fitting}\label{sec_rec}

The tracking starts with reconstructing the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the signal time and the pad position. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. We fit the chains with a helix if they have a minimum of 
10 hits. We then eliminate from the chain the hits that are 5 mm or farther from the fit,
as they are not likely part of the same track.
This new reduced chain is used for a second and final helix fit.

@@

\subsection{Resolution}

Elastic scattering was used to estimate the tracking resolution of the RTPC 
based on the residual between the expected and measured $^4$He tracks.  The RTPC
resolutions, after removing contributions from the electron, are shown in Table
~\ref{tab:reso}, and are very similar for the two halves of the RTPC.  Note that the
$\theta$- and $z$-resolutions are highly correlated.
\begin{table}[htbp]
\begin{center}
\begin{tabular}{|l|cccc|}
  \hline
& $\sigma_{z}$ &  $\sigma_{\theta}$ & $\sigma_{\phi}$ & $\sigma_{p}$\\
\hline
Left &  5.3 mm & 3.8$^{\circ}$ & 1.9$^{\circ}$ & 9$\%$ \\
Right & 6.5 mm & 4.0$^{\circ}$ & 1.9$^{\circ}$ & 8$\%$\\
\hline
\end{tabular}
\caption{The resolutions of the two modules of the RTPC for $z$-vertex, polar and azimuthal angles, and momentum.}
\label{tab:reso}
\end{center}
\end{table}

\subsection{Efficiency}

We measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two halves of the detector. We observe that the left and 
the right modules have similar efficiencies except near the upstream target 
window. This difference is due to the large number of dead channels concentrated
in this part of the left half of the detector
